A Review of Mortality Models

### 2020-09-09 mortality-models/draft.tex
WIP 9-Sep-2020

\documentclass{article}
\usepackage{amsmath, amssymb, amsthm, amsfonts}
\numberwithin{equation}{section}
\DeclareMathOperator{\pr}{P}
\linespread{1.3}
\title{A Review of Mortality Models}
\author{To be filled}
\date{Sep-2020}
\begin{document}
\maketitle
\section{Introduction}\label{s0}
Although it is very difficult to predict the behaviour of a member of a large
population one can often make statistical statements about the same population 
with much greater 
confidence. For example, it is impossible to predict when a particular 
radioactive atom will decay but one can make an accurate prediction of how 
many atoms will decay in a long enough time interval. Likewise, it is not
possible to predict the time of death of an individual although one can reliably
predict how many people from a large enough group will survive past a certain
date in future. Actuaries knew this statistical fact long before the 
mathematical foundations of the theory of probability were laid. The 
mathematical models they to describe how deaths occur in a given society are
called `mortality models'.

@@

However, the advancement in health science since the early years of the 
twentieth century, the relative peace in the western world since the second 
world war and abundance of food in the rich world resulted in a change in
historical mortality patterns. Pension funds started realizing that annuities
were being drawn for longer than expected periods. As the population in the
developed countries started greying the pension funds had the additional 
problem of a higher annunity obligations and a lower contribution from newer
workers. Pension funds now have a keen interest in understanding how mortality 
rates will evolve in the future. They now want to forecast mortality than just
produce models of the current mortality.

@@

Mortality models, being mathematical functions, produce a single curve for
a given set of its parameters. On the other hand, mortality forecasts are a
collection of curves or a family of probablity distributions. Their output
is accompanied with appropriate confidence intervals to indicate the 
uncertainty in their reported values.

Before we start reviewing the deterministic and stochastic mortality models
let us describe certain characteristic features of human mortality. Although
the life expectancy at birth has significantly improved in the post second
world war years the maximum life span seems to be pretty constant throughout 
history. A new born child faces a great risk of dying as it adjusts itself
to its new environment, warding off pathogens and fighting diseases. As time 
goes by the child starts developing immunity against many germs and its
chances of dying begin to drop. This drop continues until early adulthood 
when there is a sudden jump in mortality. It happens because of aggressive
behaviour in the late teens in both sexes and because of maternal mortality
in females. The spike in mortality eventually dwindles into a monotonic
increase in mortality as senescence catches up. We will see that there is 
a rich collection of mathematical models describing these characteristics
of human mortality. The basic structure of the mortality behaviour also
suggests that it might be profitable to fit a different model for different
stages of human life. The earliest models indeed did not cover childhood or
take into account the spike in mortality starting late teens. They were 
simple mathematical functions that fit a portion of the mortality curve that
was of most relevance to the insurance companies. Later models continued to
be deterministic in nature but had more number of parameters that could be
fitted for a given empirical data. However, it was not always possible to 
ascribe an unambiguous biological meaning to all parameters. Further, their
complexity was also controlled by the limitations of the computing technology
of their period.

We will sketch the historical development of mortality model, both 
deterministic and stochastic, starting from de Moivre's model in the next
section. We also define basic terms commonly used in actuarial science and 
demonstrate how to evaluate their expressions in simple cases. In section 3
we describe the Gompertz's model, the first realistic mortality model that 
was used for a very long time in the industry. The success of Gompertz model
lay in the number of successors it produced. Of these, we cover Makekam's 
model and its generalisations in section 4. In section 5 we mention a few
other mortality models used in European countries but whose significance today
lies mostly in their historical value. The $8$-parameter Heligman-Pollard model
was the first one that could be fitted to the entire mortality curve, starting
from infancy to very advanced years. We describe it in section 6. Starting
section 7 we turn our attention to stochastic models, starting with the
celebrated Lee-Carter model. An interesting class of models used across social
sciences is the Age-Period-Cohort models. We review a few of them relevant
to mortality in section 8. In section 9 we consider the Cairns-Blake-Dowd model
and finally we consider stochastic models with cohort effects in section 10.

\section{de Moivre's law}\label{s1}
An annuity is a lifelong payment made to an individual by a company. The 
company would want to estimate how much annuity its customers will consume
before they cease to exist. In order to estimate its yearly obligations, the
company would need to know what proportion of its customers will survive
past various points in the future. It to this problem that de Moivre 
\cite{de1731annuities} addresses his treatise. In order to estimate the value
of annuities one must consider the interest earned by the company's assets and
the probability of survival of its customers. In de Moivre's days the interest
rates were controlled by law. Therefore, he focused on estimating the 
probability of survival of the company's customers.

Developed in the 18th century, de Moivre's model was indeed quite simple. 
He assumed that there is a certain age, $\omega$ beyond which no human 
survives. Further, as time goes by the probability of a person's survival
decreases. The probability that a new born person will survive at least
$x$ years was proposed to be
\begin{equation}\label{s1e1}
s(x) = 1 - \frac{x}{\omega}, 
\end{equation}
where $0 \le x \le \omega$. The function $s$, called the survival function,
has the following characteristics
\begin{enumerate}
\item It is a decreasing function of $x$.
\item $s(x = 0) = 1$ and $s(x = \omega) = 0$. In between these age limits, it
is a linear function of $x$.
\end{enumerate}
de Moivre proposed the use of different $\omega$'s for different age groups, 
making his model piece-wise linear. We will examine a few immediate 
consequences of de Moivre's model and while doing so we will introduce the
notation common in actuarial literature \cite{jordan1967society}. The symbol
$(x)$ denotes life aged $x$. If $k$ is an estimate of the size of a population
born together then the number of people alive at age $x$ is $l_x = ks(x)$.
The number of people who will die in the interval $x$ to $x + 1$ is $d_x = l_x
- l_{x+1}$. The probability that $x$ will survive for another $n$ years is
\begin{equation}\label{s1e2}
{}_np_x = \frac{l_{x+n}}{l_x} = \frac{s(x+n)}{s(x)}.
\end{equation}
The probability that a person will die within next $n$ years of $x$ is
\begin{equation}\label{s1e3}
{}_nq_x = 1 - {}_np_x.
\end{equation}
It is common to denote ${}_1p_x$ as $p_x$ and ${}_nq_x$ as $q_x$. The `force
of mortality' is defined as
\begin{equation}\label{s1e4}
\mu(x) = -\frac{s^\prime(x)}{s(x)}, 
\end{equation}
where a $s^\prime$ denotes the derivative of $s$ with respect to $x$. The 
force of mortality is also called the hazard function in the statistical
literature. The hazard function or the `force of mortality' is an individual's
susceptibility to die. Its reciprocal is the individual's resistance to death. 
de Moivre's law proposes that a person's resistance to die varies linearly over
time. That is,
\begin{equation}\label{s1e5}
\nu(x) = \frac{1}{\mu(x)} = \omega - x.
\end{equation}
If $x_0 = a, x_1 = a + d, x_2 = a + 2d, ..., x_n = a + nd$ then it is clear 
that $\nu(x_n)$ form an algebraic progression. This point may seem to be just a 
mathematical characteristic at this stage. We shall see in the next section that
it marked a departure from de Moivre's model to a more realistic one. For de 
Moivre's law of mortality, it is easy to confirm that
\begin{eqnarray}
{}_np_x &=& \frac{\omega - (x + n)}{\omega - x} \label{s1e6} \\
{}_nq_x &=& \frac{n}{\omega - x} \label{s1e7} \\
\mu(x) &=& \frac{1}{\omega - x} \label{s1e8}
\end{eqnarray}
de Moivre himself did not consider his law to be an accurate description of
human mortality. He emphasized in his treatise that it was an approximation
useful to calculate the annuities. His piecewise linear approximation was a
valuable device to carry out practical computations in the eighteenth century.

\section{Gompertz's law}\label{s2}
An immediate consequence of de Moivre's law is that an individual's resistance
to death diminishes algebraically as time goes by. The English actuary 
Benjamin Gompertz observed \cite{gompertz1825xxiv} that the resistance to 
death decreases geometrically instead. If one considers small enough time 
intervals then at the end of every interval a person's resistance to death 
reduces by the same proportion. Expressed mathematically, it means that
\begin{equation}\label{s2e1}
\frac{d\nu}{dx} = -k\nu(x),
\end{equation}
where the function $\nu$ was defined in equation \eqref{s1e5}. It immediately
follows that
\begin{equation}\label{s2e2}
\log\nu(x) = -kx + \log B,
\end{equation}
where $B$ is a constant of integration. The `force of mortality' or the
hazard function is therefore
\begin{equation}\label{s2e3}
\mu(x) = Be^{kx}.
\end{equation}
The constant $B$ is called baseline mortality and $k$ is called the 
senescent component.
An individual's susceptibility to die thus increases exponentially with age.
From equation \eqref{s1e4} and \eqref{s2e3}, the differential equation for
the survival function is
\begin{equation}\label{s2e4}
\frac{s^\prime(x)}{s(x)} = Be^{kx}.
\end{equation}
The solution of this ordinary differential equation is
\begin{equation}\label{s2e5}
s(x) = C\exp\left(\frac{B}{k}\exp(kx)\right),
\end{equation}
where $C$ is another constant of integration. The constants $B$ and $C$ in
equation \eqref{s2e5} are determined from the mortality data of the population
under study. Gompertz suggested that his formula is applicable for the age
group between $10-15$ years of age up to $55-60$. In order to extend it beyond
the age of $60$, Gompertz recommended that a different set of constants $B$
and $C$ be used.

Gompertz was aware that death in a population has at least two factors. One
of them is the deterioration in the individual's resistance to death. The 
other one is chance alone. The first factor depends on the individual's age
while the second factor is constant throughout. However, in the formulation
of his law of mortality, Gompertz considered only the first factor.

\section{Makeham's law and its generalisations}\label{s3}
William Makeham, the English mathematician and actuary, added a term to
Gompertz's law of resistance to death which the latter had acknowledged but
not included. Makeham's modification \cite{makeham1867law} of equation 
\eqref{s2e3} is
\begin{equation}\label{s3e1}
\mu(x) = A + Be^{kx},
\end{equation}
where $A$ is a constant independent of the person's age. Makeham's survival
function is
\begin{equation}\label{s3e2}
s(x) = C\exp\left(Ax + \frac{B}{k}\exp(kx)\right).
\end{equation}
Makeham's survival function has three constants $A, B$ and $C$ and therefore
can be fitted to emprically observed mortality data more effectively than
Gompertz's model. Makeham's law was often applied to mortality tables in the
age range beyond $20$ years to almost end of life\cite{jordan1967society}. 
In 1890\cite{makeham1890further}, Makeham proposed yet another modification to 
his law by adding a linear term to the hazard function or the force of 
mortality. Makeham thought that in addition to the two factors spotted by
Gompertz there is another one that grows linearly with age. It was proposed to 
be
\begin{equation}\label{s3e3}
\mu(x) = A_1 + A_2x + Be^{kx}.
\end{equation}
The resulting survival function is
\begin{equation}\label{s3e4}
s(x) = C\exp\left(A_1x + \frac{A_2}{2}x^2 + \frac{B}{k}\exp(kx)\right).
\end{equation}
Makeham's law, like its predecessor, Gompertz's law was used to construct
mortality tables of adults. Their simplicity allowed them to be used 
effectively up to as late as the 1940s. Gompertz law was used to construct
the 1937 Standard Annuity Table and Makeham's law for 1940 Annuity table
\cite{jordan1967society}. Both laws assumed that the force of mortality 
increases monotonically with time. However, it is commonly observed that 
infant mortality contradicts this assumption. In order to extend Makeham's
law of 1860 to infantile ages, Wilhem Lazarus \cite{lazarus1867ueber} added
an exponentially decreasing term. His law expressed the force of mortality
as 
\begin{equation}\label{s3e5}
\mu(x) = B_1 e^{-k_1x} + A + B_2e^{k_2 x}.
\end{equation}
The constants $B_1$ and $B_2$ are positive numbers. The corresponding survival 
function is
\begin{equation}\label{s3e6}
s(x) = 
C\exp\left(-\frac{B_1}{k_1}e^{-k_1x} + Ax + \frac{B_2}{k_2}e^{k_2x}\right)
\end{equation}

The way infants' susceptibility to die of diseases upsets Gompertz's and 
Makeham's assumption of monotonicity of the law of mortality, do does the 
spike in the death rate of young adults. Wars were frequent in Europe of the 
eighteenth century and the victims were disproportionately the young recruits.
Even otherwise, young adults are more prone to death by accidents. In order
to take into account this mortality hump in the young, Thiele 
\cite{thiele1871mathematical} introduced a gaussian term to Lazarus' model.
His law of mortality is
\begin{equation}\label{s3e7}
\mu(x) = B_1 e^{-k_1x} + Ae^{-a_1(x - a_2)^2} + B_2e^{k_2 x}.
\end{equation}
All the constants in this equation, namely $A, a_1, a_2, B_1, k_1, B_2$ and 
$k_2$ are positive numbers. The presence of the gaussian term prevents us
from writing a closed form expression of the survival function. However, the
ordinary differential equation for the survival function can be readily 
integrated numerically. Alternatively, if the initial conditions are available,
one can write its solution in terms of the error function.

We observe that all generalisations of Makeham's laws introduce additional
constants to the force of mortality and the hazard function. These constants
provide additional `degrees of freedom' to the functions used to fit the 
mortality data, resulting in a better fit.

Forfar, McCutcheon and Wilkie proposed a far wider generalisation of the
Gompertz and Makeham's laws. They defined a class of models, called GM models
after their inventors, with the mathematical form
\begin{equation}\label{s3e8}
\mu(x) = \sum_{i=0}^{r-1}\alpha_i x^i + 
\exp\left(\sum_{i=1}^{s-1}\beta_i x^i\right).
\end{equation}
They also used the convention that if $r = 0$ then the polynomial terms are 
absent and if $s = 0$ then the exponential terms do not appear. They called
the law of force of mortality of equation \eqref{s3e8} as $\mathrm{GM}(r,s)$.
It is easy to check that Gompertz's law is $\mathrm{GM}(0, 2)$, Makeham's
first law is $\mathrm{GM}(1,2)$ and Makeham's second law is $\mathrm{GM}(2,2)$.
The $\mathrm{GM}(0, 2)$, $\mathrm{GM}(2,2)$ and $\mathrm{GM}(1,3)$ are used
by the Continuous Mortality Investigation Bureau\cite{pitacco2016high} in the 
United Kingdom.

\section{Other mortality laws prior to the 20th century}\label{s4}
de Moivre's, Gompertz's and Makeham's laws of mortality were not the only
ones in vogue in the nineteenth century and prior to it. In this section we
mention a few notable ones without commenting on their merits. Lambert's
survival function \cite{lambert1776dottrina} was
\begin{equation}\label{s4e1}
s(x) = \left(\frac{a - x}{x}\right)^2 - b\left(e^{-x/c} - e^{-x/d}\right).
\end{equation}
It had four degrees of freedom. Babbage\cite{babbage1823tables} proposed a 
quadratic survival function
\begin{equation}\label{s4e2}
s(x) = c - bx - ax^2.
\end{equation}
Young proposed a polynomial of degree $40$ as the survival function in 
\cite{adler1866memoir} 1826. von Littrow\cite{von1832lebensversicherungen}
extended Babbage's model to polynomials of higher degree. Moser
\cite{moser1839gesetze} proposed a polynomial survival function with 
fractional powers of $x$. It has five degrees of freedom and had the form
\begin{equation}\label{s4e3}
s(x) = 1 - ax^{1/4} + bx^{9/4} - cx^{17/4} - dx^{25/4} + ex^{33/4}.
\end{equation}
All these models were attempts to fit the mortality data with a continuous 
function. They were not rooted in demographic observations like the models
of Gompertz, Makeham, Lazarus or Thiele. As a result, they soon fell into
disuse after Gompertz proposed his model and others improved upon it.

\section{Heligman-Pollard model}\label{s5}
The term `odds' is a well-defined term in probability theory. When we are 
dealing with an experiment with two outcomes - that something happens or 
it does not - then it is common in certain fields to express the 
probabilities in terms of odds.  For example, if $p$ is the probability 
that an event will happen and $q = 1 - p$ is the probability that it will 
not happen then the odds of the experiment is $p/q$.

@@

The advantages of Lee-Carter model lie in its robustness, simplicity and 
the fact that it fits well over wide ranges of age. It is also observed to be
able to fit small non-linearities in the mortality curve quite well. Its 
shortcomings are its lack of smoothness of age effects in small populations, 
inability to work with cohorts, inability to cope with improvements varying 
across times and a possible underestimation of uncertainty. It also has a 
limited versatality owing to it being a one-factor model.

>>>>

\documentclass{article}
\usepackage{amsmath, amssymb, amsthm, amsfonts}
\numberwithin{equation}{section}
\DeclareMathOperator{\pr}{P}
\linespread{1.3}
\title{A Review of Mortality Models}
\author{To be filled}
\date{Sep-2020}
\begin{document}
\maketitle
\section{Introduction}\label{s0}
Although it is very difficult to predict the behavior of a member of a large
population one can often make statistical statements about the same population 
with much greater 
confidence. For example, it is impossible to predict when a particular 
radioactive atom will decay but one can make an accurate prediction of how 
many atoms will decay in a long enough time interval. Likewise, it is not
possible to predict the time of death of an individual although one can reliably
predict how many people from a large enough group will survive past a certain
date in future. Actuaries knew this statistical fact long before the 
mathematical foundations of the theory of probability were laid. The 
mathematical models they to describe how deaths occur in a given society are
called `mortality models'.

@@

However, the advancement in health science since the early years of the 
twentieth century, the relative peace in the western world since the second 
world war and abundance of food in the rich world resulted in a change in
historical mortality patterns. Pension funds started realizing that annuities
were being drawn for longer than expected periods. As the population in the
developed countries started graying the pension funds had the additional 
problem of a higher annuity obligations and a lower contribution from newer
workers. Pension funds now have a keen interest in understanding how mortality 
rates will evolve in the future. They now want to forecast mortality than just
produce models of the current mortality.

@@

Mortality models, being mathematical functions, produce a single curve for
a given set of its parameters. On the other hand, mortality forecasts are a
collection of curves or a family of probability distributions. Their output
is accompanied with appropriate confidence intervals to indicate the 
uncertainty in their reported values.

Before we start reviewing the deterministic and stochastic mortality models
let us describe certain characteristic features of human mortality. Although
the life expectancy at birth has significantly improved in the post second
world war years the maximum life span seems to be pretty constant throughout 
history. A new born child faces a great risk of dying as it adjusts itself
to its new environment, warding off pathogens and fighting diseases. As time 
goes by the child starts developing immunity against many germs and its
chances of dying begin to drop. This drop continues until early adulthood 
when there is a sudden jump in mortality. It happens because of aggressive
behavior in the late teens in both sexes and because of maternal mortality
in females. The spike in mortality eventually dwindles into a monotonic
increase in mortality as senescence catches up. We will see that there is 
a rich collection of mathematical models describing these characteristics
of human mortality. The basic structure of the mortality behavior also
suggests that it might be profitable to fit a different model for different
stages of human life. The earliest models indeed did not cover childhood or
take into account the spike in mortality starting late teens. They were 
simple mathematical functions that fit a portion of the mortality curve that
was of most relevance to the insurance companies. Later models continued to
be deterministic in nature but had more number of parameters that could be
fitted for a given empirical data. However, it was not always possible to 
ascribe an unambiguous biological meaning to all parameters. Further, their
complexity was also controlled by the limitations of the computing technology
of their period.

We will sketch the historical development of mortality model, both 
deterministic and stochastic, starting from de Moivre's model in the next
section. We also define basic terms commonly used in actuarial science and 
demonstrate how to evaluate their expressions in simple cases. In section 3
we describe the Gompertz's model, the first realistic mortality model that 
was used for a very long time in the industry. The success of Gompertz model
lay in the number of successors it produced. Of these, we cover Makekam's 
model and its generalizations in section 4. In section 5 we mention a few
other mortality models used in European countries but whose significance today
lies mostly in their historical value. The $8$-parameter Heligman-Pollard model
was the first one that could be fitted to the entire mortality curve, starting
from infancy to very advanced years. We describe it in section 6. Starting
section 7 we turn our attention to stochastic models, starting with the
celebrated Lee-Carter model. An interesting class of models used across social
sciences is the Age-Period-Cohort models. We review a few of them relevant
to mortality in section 8. In section 9 we consider the Cairns-Blake-Dowd model
and finally we consider stochastic models with cohort effects in section 10.

\section{de Moivre's law}\label{s1}
An annuity is a lifelong payment made to an individual by a company. The 
company would want to estimate how much annuity its customers will consume
before they cease to exist. In order to estimate its yearly obligations, the
company would need to know what proportion of its customers will survive
past various points in time in the future. It to this problem that de Moivre 
\cite{de1731annuities} addressed his treatise. In order to estimate the value
of annuities one must consider the interest earned by the company's assets and
the probability of survival of its customers. In de Moivre's days the interest
rates were controlled by law. Therefore, he focused on estimating the 
probability of survival of the company's customers.

Developed in the 18th century, de Moivre's model was indeed quite simple. 
He assumed that there is a certain age, $\omega$ beyond which no human 
survives. Further, as time goes by the probability of a person's survival
decreases. The probability that a new born person will survive at least
$x$ years was proposed to be
\begin{equation}\label{s1e1}
s(x) = 1 - \frac{x}{\omega}, 
\end{equation}
where $0 \le x \le \omega$. The function $s$, called the survival function,
has the following characteristics
\begin{enumerate}
\item It is a decreasing function of $x$.
\item $s(x = 0) = 1$ and $s(x = \omega) = 0$. In between these age limits, it
is a linear function of $x$.
\end{enumerate}
de Moivre proposed the use of different $\omega$'s for different age groups, 
making his model piece-wise linear. We will examine a few immediate 
consequences of de Moivre's model and while doing so we will introduce the
notation common in actuarial literature \cite{jordan1967society}. The symbol
$(x)$ denotes life aged $x$. If $k$ is an estimate of the size of a population
born together then the number of people alive at age $x$ is $l_x = ks(x)$.
The number of people who will die in the interval $x$ to $x + 1$ is $d_x = l_x
- l_{x+1}$. The probability that $x$ will survive for another $n$ years is
\begin{equation}\label{s1e2}
{}_np_x = \frac{l_{x+n}}{l_x} = \frac{s(x+n)}{s(x)}.
\end{equation}
The probability that a person will die within next $n$ years of $x$ is
\begin{equation}\label{s1e3}
{}_nq_x = 1 - {}_np_x.
\end{equation}
It is common to denote ${}_1p_x$ as $p_x$ and ${}_nq_x$ as $q_x$. The `force
of mortality' is defined as
\begin{equation}\label{s1e4}
\mu(x) = -\frac{s^\prime(x)}{s(x)}, 
\end{equation}
where a $s^\prime$ denotes the derivative of $s$ with respect to $x$. The 
force of mortality is also called the hazard function in the statistical
literature. The hazard function or the `force of mortality' is an individual's
susceptibility to die. Its reciprocal is the individual's resistance to death. 
de Moivre's law proposes that a person's resistance to die varies linearly over
time. That is,
\begin{equation}\label{s1e5}
\nu(x) = \frac{1}{\mu(x)} = \omega - x.
\end{equation}
If $x_0 = a, x_1 = a + d, x_2 = a + 2d, ..., x_n = a + nd$ then it is clear 
that $\nu(x_n)$ form an algebraic progression. This point may seem to be just a 
mathematical characteristic at this stage. We shall see in the next section that
it marked a departure from de Moivre's model to a more realistic one. For de 
Moivre's law of mortality, it is easy to confirm that
\begin{eqnarray}
{}_np_x &=& \frac{\omega - (x + n)}{\omega - x} \label{s1e6} \\
{}_nq_x &=& \frac{n}{\omega - x} \label{s1e7} \\
\mu(x) &=& \frac{1}{\omega - x} \label{s1e8}
\end{eqnarray}
de Moivre himself did not consider his law to be an accurate description of
human mortality. He emphasized in his treatise that it was an approximation
useful to calculate the annuities. His piece-wise linear approximation was a
valuable device to carry out practical computations in the eighteenth century.

\section{Gompertz's law}\label{s2}
An immediate consequence of de Moivre's law is that an individual's resistance
to death diminishes algebraically as time goes by. The English actuary 
Benjamin Gompertz observed \cite{gompertz1825xxiv} that the resistance to 
death decreases geometrically instead. If one considers small enough time 
intervals then at the end of every interval a person's resistance to death 
reduces by the same proportion. Expressed mathematically, it means that
\begin{equation}\label{s2e1}
\frac{d\nu}{dx} = -k\nu(x),
\end{equation}
where the function $\nu$ was defined in equation \eqref{s1e5}. It immediately
follows that
\begin{equation}\label{s2e2}
\log\nu(x) = -kx + \log B,
\end{equation}
where $B$ is a constant of integration. The `force of mortality' or the
hazard function is therefore
\begin{equation}\label{s2e3}
\mu(x) = Be^{kx}.
\end{equation}
The constant $B$ is called baseline mortality and $k$ is called the 
senescent component.
An individual's susceptibility to die thus increases exponentially with age.
From equation \eqref{s1e4} and \eqref{s2e3}, the differential equation for
the survival function is
\begin{equation}\label{s2e4}
\frac{s^\prime(x)}{s(x)} = Be^{kx}.
\end{equation}
The solution of this ordinary differential equation is
\begin{equation}\label{s2e5}
s(x) = C\exp\left(\frac{B}{k}\exp(kx)\right),
\end{equation}
where $C$ is another constant of integration. The constants $B$ and $C$ in
equation \eqref{s2e5} are determined from the mortality data of the population
under study. It is common to introduce a constant $c = e^k$ and write equation
\eqref{s2e5} as
\begin{equation}\label{s2e6}
s(x) = C\exp\left(\frac{B}{k}c^x\right),
\end{equation}
Gompertz suggested that his formula is applicable for the age
group between $10-15$ years of age up to $55-60$. In order to extend it beyond
the age of $60$, Gompertz recommended that a different set of constants $B$
and $C$ be used.

Gompertz was aware that death in a population has at least two factors. One
of them is the deterioration in the individual's resistance to death. The 
other one is chance alone. The first factor depends on the individual's age
while the second factor is constant throughout. However, in the formulation
of his law of mortality, Gompertz considered only the first factor. Gompertz
could have fitted simpler polynomial functions to the mortality data available
in his times. However, instead of trying to fit a function to the data he
developed his model based on an observation root in biology that organisms
deteriorate with age in a geometric proportion. Gompertz model was not only a
better fit to data than de Moivre's but it also started the tradition of
finding a biological explanation for every term in a mortality model. It is
important to note that Gompertz considered death as just a biological 
phenomenon. His law is best applicable to populations well protected from 
wars or pestilence. It is, therefore, an upper limit to longevity in a
population. In that sense, it served to be a useful estimate of the insurance
companies' liabilities. The success of Gompertz's law in practical calculations
ensured that for several decades newer mortality laws were only an improvement
over it. In the next section, we shall consider a few of them.

\section{Makeham's law and its generalizations}\label{s3}
William Makeham, the English mathematician and actuary, added a term to
Gompertz's law of resistance to death which the latter had acknowledged but
not included. Makeham's modification \cite{makeham1867law} of equation 
\eqref{s2e3} is
\begin{equation}\label{s3e1}
\mu(x) = A + Be^{kx},
\end{equation}
where $A$ is a constant independent of the person's age. Makeham's survival
function is
\begin{equation}\label{s3e2}
s(x) = C\exp\left(Ax + \frac{B}{k}\exp(kx)\right).
\end{equation}
Makeham's survival function has three constants $A, B$ and $C$ and therefore
can be fitted to empirically observed mortality data more effectively than
Gompertz's model. If we write $c = e^k$ then we can write equation \eqref{s3e2}
in its customary form
\[
s(x) = C\exp\left(Ax + \frac{B}{k}c^x\right).
\]
Makeham's law was often applied to mortality tables in the
age range beyond $20$ years to almost end of life\cite{jordan1967society}. 
In 1890\cite{makeham1890further}, Makeham proposed yet another modification to 
his law by adding a linear term to the hazard function or the force of 
mortality. Makeham thought that in addition to the two factors spotted by
Gompertz there is another one that grows linearly with age. It was proposed to 
be
\begin{equation}\label{s3e3}
\mu(x) = A_1 + A_2x + Be^{kx}.
\end{equation}
The resulting survival function is
\begin{equation}\label{s3e4}
s(x) = C\exp\left(A_1x + \frac{A_2}{2}x^2 + \frac{B}{k}\exp(kx)\right).
\end{equation}
Makeham's law, like its predecessor, Gompertz's law was used to construct
mortality tables of adults. Their simplicity allowed them to be used 
effectively up to as late as the 1940s. Gompertz law was used to construct
the 1937 Standard Annuity Table and Makeham's law for 1940 Annuity table
\cite{jordan1967society}. Both laws assumed that the force of mortality 
increases monotonically with time. However, it is commonly observed that 
infant mortality contradicts this assumption. This is the reason why Gompertz's
and Makeham's laws were applied only to populations over $20$ years of age.

Mortality decreases from the birth of the child until the onset of adulthood. 
In order to extend Makeham's law of 1860 to infantile ages, Wilhelm Lazarus 
\cite{lazarus1867ueber} added an exponentially decreasing term. His law 
expressed the force of mortality as 
\begin{equation}\label{s3e5}
\mu(x) = B_1 e^{-k_1x} + A + B_2e^{k_2 x}.
\end{equation}
The constants $B_1$ and $B_2$ are positive numbers. The corresponding survival 
function is
\begin{equation}\label{s3e6}
s(x) = 
C\exp\left(-\frac{B_1}{k_1}e^{-k_1x} + Ax + \frac{B_2}{k_2}e^{k_2x}\right)
\end{equation}

The way infants' susceptibility to die of diseases upsets Gompertz's and 
Makeham's assumption of monotonicity of the law of mortality, so does the 
spike in the death rate of young adults. Wars were frequent in Europe of the 
eighteenth century and the victims were disproportionately the young recruits.
Even otherwise, young adults are more prone to death by accidents. Maternal
mortality too was quite high in the eighteenth century. In order
to take into account this mortality hump in the young, Thiele 
\cite{thiele1871mathematical} introduced a Gaussian term to Lazarus' model.
His law of mortality is
\begin{equation}\label{s3e7}
\mu(x) = B_1 e^{-k_1x} + Ae^{-a_1(x - a_2)^2} + B_2e^{k_2 x}.
\end{equation}
All the constants in this equation, namely $A, a_1, a_2, B_1, k_1, B_2$ and 
$k_2$ are positive numbers. The presence of the Gaussian term prevents us
from writing a closed form expression of the survival function. However, the
ordinary differential equation for the survival function can be readily 
integrated numerically. Alternatively, if the initial conditions are available,
one can write its solution in terms of the error function.

We observe that all generalizations of Makeham's laws introduce additional
constants to the force of mortality and the hazard function. These constants
provide additional `degrees of freedom' to the functions used to fit the 
mortality data, resulting in a better fit.

Forfar, McCutcheon and Wilkie \cite{forfar1988graduation} proposed a far wider 
generalization of the Gompertz and Makeham's laws. They defined a class of 
models, called GM models after their inventors, with the mathematical form
\begin{equation}\label{s3e8}
\mu(x) = \sum_{i=0}^{r-1}\alpha_i x^i + 
\exp\left(\sum_{i=1}^{s-1}\beta_i x^i\right).
\end{equation}
They also used the convention that if $r = 0$ then the polynomial terms are 
absent and if $s = 0$ then the exponential terms do not appear. They called
the law of force of mortality of equation \eqref{s3e8} as $\mathrm{GM}(r,s)$.
It is easy to check that Gompertz's law is $\mathrm{GM}(0, 2)$, Makeham's
first law is $\mathrm{GM}(1,2)$ and Makeham's second law is $\mathrm{GM}(2,2)$.
The $\mathrm{GM}(0, 2)$, $\mathrm{GM}(2,2)$ and $\mathrm{GM}(1,3)$ are used
by the Continuous Mortality Investigation Bureau\cite{pitacco2016high} in the 
United Kingdom.

Gompertz's model and its descendants considered in this section had simple
algebraic forms, each term of which could be interpreted in biological
terms. However, they were not always easy to fit on available mortality
data. The presence of exponentials also made them numerically unstable. 
Moreover, they covered only a limited number of sources of mortality.

\section{Other deterministic mortality laws}\label{s4}
de Moivre's, Gompertz's and Makeham's laws of mortality were not the only
ones in vogue in the nineteenth century and prior to it. In this section we
mention a few notable ones without commenting on their merits. Lambert's
survival function \cite{lambert1776dottrina} was
\begin{equation}\label{s4e1}
s(x) = \left(\frac{a - x}{x}\right)^2 - b\left(e^{-x/c} - e^{-x/d}\right).
\end{equation}
It had four degrees of freedom. Babbage\cite{babbage1823tables} proposed a 
quadratic survival function
\begin{equation}\label{s4e2}
s(x) = c - bx - ax^2.
\end{equation}
Young proposed a polynomial of degree $40$ as the survival function in 
\cite{adler1866memoir} 1826. von Littrow\cite{von1832lebensversicherungen}
extended Babbage's model to polynomials of higher degree. Moser
\cite{moser1839gesetze} proposed a polynomial survival function with 
fractional powers of $x$. It has five degrees of freedom and had the form
\begin{equation}\label{s4e3}
s(x) = 1 - ax^{1/4} + bx^{9/4} - cx^{17/4} - dx^{25/4} + ex^{33/4}.
\end{equation}
The original mortality laws of Gompertz and Makeham were applicable only
for adults. Oppermann proposed \cite{oppermann1870graduation} a law for people 
younger than $20$ years of age. It was
\begin{equation}\label{s4e4}
\mu(x) = \frac{a}{\sqrt{x}} + b + cx^{1/3},
\end{equation}
where $a, b, c$ are parameter selected to fit the data. Perks 
\cite{perks1932some} proposed two models of the form
\begin{eqnarray}
\mu_1(x) &=& \frac{A + Bc^x}{1 + Dc^x} \label{s4e5} \\
\mu_2(x) &=& \frac{A + Bc^x}{Kc^{-x} + 1 + Dc^x} \label{s4e6}
\end{eqnarray}
The denominators in these forms flatten the exponential increase in Gompertz
form, especially at advanced age. The mortality law of equation \eqref{s4e5}
was used to graduate table of annuitant lives in the United Kingdom. Beard
\cite{beard1971some} wrote his law in terms of the probability that a life 
aged $x$ will die in the next year. It was
\begin{equation}\label{s4e7}
q_x = A + \frac{Bc^x}{Ex^{-2x} + 1 + Dc^x},
\end{equation}
where $A, B, c, D, E$ are parameters selected to fit the mortality data.
Barnett's formula appears on page 141 of \cite{joint1974considerations}. It
fits data using four parameters and has the mathematical form
\begin{equation}\label{s4e8}
q_x = \frac{A - Hx - Bc^x}{1 + A - Hx + Bc^x}.
\end{equation}
Wilkie \cite{wilkie1976international} generalized this form to
\begin{equation}\label{s4e9}
q_x = \frac{f(x)}{1 + f(x)},
\end{equation}
where
\begin{equation}\label{s4e10}
f(x) = \exp\left(\sum_{i=1}^s\alpha_i x^{i-1}\right).
\end{equation}
All these models were attempts to fit the mortality data with a continuous 
function. Many were increasingly complicated functions that were selected on
the sole merit that they fitted the observed data well. They were not rooted in 
demographic observations like the models of Gompertz, Makeham, Lazarus or 
Thiele. As a result, they soon fell into disuse after stochastic models came on
the scene.

\section{Heligman-Pollard model}\label{s5}
The term `odds' is a well-defined term in probability theory. When we are 
dealing with an experiment with two outcomes - that something happens or 
it does not - then it is common to express the probabilities in terms of odds. 
For example, if $p$ is the probability that an event will happen and $q = 1 - p$
is the probability that it will not happen then the odds of the experiment is 
$p/q$. The usage of odds instead of probabilities is very common in gambling or
betting.

@@

The advantages of Lee-Carter model lie in its robustness, simplicity and 
the fact that it fits well over wide ranges of age. It is also observed to be
able to fit small non-linearities in the mortality curve quite well. Its 
shortcomings are its lack of smoothness of age effects in small populations, 
inability to work with cohorts, inability to cope with improvements varying 
across times and a possible underestimation of uncertainty. It also has a 
limited versatility owing to it being a one-factor model.
